The State of Urbit:\\Eight Years After the Whitepaper

### 2024-04-25 mss.tex
Final cut.

While the whitepaper by its own admission presented itself as a ``work-in-progress presentation'' of a ``semi-closed alpha test'', the basic claims of the Urbit whitepaper are still true today.  Most of the intervening work over the past eight years has gone into improving robustness and better support for applications.  A capsule history of ``late early'' Urbit includes the following milestones:
\begin{itemize}
  \item  2015–2017:  Early vane explosion
  \item  2017–2019:  Hoon, Ford, Eyre, and Vere rewrites refine the system
  \item  2018–2019:  Azimuth and Ames rewrite refine the network stack
  \item  2018–2019:  Nock bytecode interpreter
  \item  2020:  Ford Fusion enables network continuity
  \item  2020:  Security audit on Ames
  \item  2021:  Azimuth Layer2
  \item  2021:  Software distribution introduced
  \item  2022–2024:  Software distribution maturation
  \item  2023–2024:  Network scaling
\end{itemize}

@@

\sloppy
Going on-chain also enabled ``personal continuity breaches'', in which a ship owner can notify the network through the chain that their Urbit node was rebooted from scratch, due to data loss, unrecoverable error, or compromised security.  This feature necessitated some rework of the kernel to ensure data consistency across vanes when receiving a notification that a peer had breached.

@@

\subsection{Essential Desks (2023/2024)}

@@

The Ares project, now close to initial deployment, includes a clean-slate rewrite of the Nock interpreter, using an all-new allocator, a novel static analysis technique called Subject Knowledge Analysis to achieve substantial Nock execution speed improvements, and a new persistence module called the Persistent Memory Arena, using a copy-on-write B+ tree to keep memory pages synchronized between disk and {\sc ram}.

@@

Fortunately, because Urbit supports structural sharing and Ford is a purely functional build system, a referentially transparent cache was added to Ford.  This cache does not know which desk a build is being run in, so it can cache and deduplicate builds across desks seamlessly without losing the logical static linking (see pp. 75–82 in this issue).

@@

Urbit development has begun to realize many of the promises of the 2016 whitepaper.  The system is more reliable, more scalable, more extensible, and more secure than it was then.  The network has grown to include thousands of users, and the system is now capable of supporting a much larger userbase.  Indeed, this article omits many quality-of-life improvements, such as scrying over {\sc http}, EAuth, \lstinline[style=inlinecode]{%bout} timing hints, bootstrap pill management, reliability and usability of Tlon's social media application, color printing, ordered map data structures, compile-time evaluation in Hoon, the Bridge contract operations tool for Azimuth and details of the Urbit HD wallet, the 
The published roadmap (available at \url{https://roadmap.urbit.org/}) provides more detail on past and planned accomplishments at a resolution impossible in a single article.

@@

The \lstinline[style=inlinecode]{.^} dotket rune no longer blocks until complete.  Because the agent model expects non-blocking I/O, a single dotket could make an entire agent stuck if it can't be completed.  While a lot of work has gone into fleshing out the scry namespace, core developers are now entertaining the notion that the expectation of synchronous access to the namespace might not be worth it for agents.  An Urbit Improvement Proposal has been submitted that would remove the \lstinline[style=inlinecode]{.^} dotket rune in favor of a move-based interface to the namespace (see pp. 47–58 in this issue).

>>>>

While the whitepaper by its own admission presented itself as a ``work-in-progress presentation'' of a ``semi-closed alpha test'', the basic claims of the Urbit whitepaper are still true today.  Most of the intervening work over the past eight years has gone into improving robustness and better support for applications.  A capsule history of ``late early'' Urbit includes the following milestones:
\begin{itemize}
  \item  2015–2017:  Early vane explosion
  \item  2017–2019:  Rewrites of Hoon, Ford, Eyre, and Vere refine the system
  \item  2018–2019:  Rewrites of Azimuth and Ames refine the network stack
  \item  2018–2019:  Nock bytecode interpreter
  \item  2020:  Ford Fusion enables network continuity
  \item  2020:  Security audit on Ames
  \item  2021:  Azimuth Layer2
  \item  2021:  Software distribution introduced
  \item  2022–2024:  Software distribution maturation
  \item  2023–2024:  Network scaling
\end{itemize}

@@

\sloppy
Going on-chain also enabled ``personal continuity breaches'', in which a ship (instance) owner can notify the network through the chain that their Urbit node was rebooted from scratch, due to data loss, unrecoverable error, or compromised security.  This feature necessitated some rework of the kernel to ensure data consistency across vanes when receiving a notification that a peer had breached.

@@

\subsection{Essential Desks (2023–24)}

@@

The Ares project, close to initial deployment as of writing, includes a clean-slate rewrite of the Nock interpreter, using an all-new allocator, a novel static analysis technique called Subject Knowledge Analysis to achieve substantial Nock execution speed improvements, and a new persistence module called the Persistent Memory Arena, using a copy-on-write B+ tree to keep memory pages synchronized between disk and {\sc ram}.

@@

Fortunately, because Urbit supports structural sharing and Ford is a purely functional build system, a referentially transparent cache was added to Ford.  This cache does not know which desk a build is being run in, so it can cache and deduplicate builds across desks seamlessly without losing the logical static linking (see \citet{Blackman2024}, pp. 75–82 in this issue).

@@

Urbit development has begun to realize many of the promises of the 2016 whitepaper.  The system is more reliable, more scalable, more extensible, and more secure than it was then.  The network has grown to include thousands of users, and the system is now capable of supporting a much larger userbase.  Indeed, this article omits many quality-of-life improvements, such as scrying over {\sc http}, EAuth, \lstinline[style=inlinecode]{%bout} timing hints, bootstrap pill management, reliability and usability of Tlon's social media application, color printing, ordered map data structures, compile-time evaluation in Hoon, the Bridge contract operations tool for Azimuth, and details of the Urbit HD wallet.  More detail on past and planned accomplishments are provided at the published roadmap, \url{https://roadmap.urbit.org/}.

@@

The \lstinline[style=inlinecode]{.^} dotket rune no longer blocks until complete.  Because the agent model expects non-blocking I/O, a single dotket could make an entire agent stuck if it can't be completed.  While a lot of work has gone into fleshing out the scry namespace, core developers are now entertaining the notion that the expectation of synchronous access to the namespace might not be worth it for agents.  An Urbit Improvement Proposal has been submitted that would remove the \lstinline[style=inlinecode]{.^} dotket rune in favor of a move-based interface to the namespace (see \citet{Blackman2024a}, pp. 47–58 in this issue).
